manuscript_v1

### 2022-02-14 manuscript.Rmd
manuscript v1.5 (Discussion complete)

Heterogeneity has been explored before in habitat selection studies, but the focus has mostly been on the habitat and not the canopy. For example, Freemark and Merriam (1986) observed positive trends of bird richness and abundance with overall habitat heterogeneity (combination of species compositions and forest structure) and forest size, which are expected to correlate. Canopy heterogeneity however, along with canopy cover, affect light availability within the forest, which would result in higher foraging efficiency and thus can be crucial in foraging for invertebrates in autumn when the conspicuous prey are depleted. Heterogeneity could also explain higher omnivore abundance in forest edges. In our study, across multiple points of the same area within the same forest tract, the relation between omnivore abundance and canopy heterogeneity evolved with time, suggesting that these preferences are driven directly or indirectly by the seasonally changing conditions. *Canopy cover was not included as a candidate predictor and hence its effect cannot be compared*, but heterogeneity might potentially be more important to birds than average cover, as heterogeneous patches provide better light conditions while still retaining some other desirable properties of the habitat. Patches with low cover, despite having good light conditions, might not have sufficiently good vegetation characteristics. This could be indicating two separable properties of the canopy that often go unnoticed.

Despite the absence of a significant correlation between canopy heterogeneity and tree density, the latter was negatively related to bird abundance (at the community level as well in omnivores). Since a vast majority of bird detections were auditory and since auditory detections are not limited by such temperate forests with poor understorey layers, the possibility of poorer detection rates with higher tree density can be ruled out, though it is unknown whether density also affects overall bird behaviour. In omnivores, the structural preference rather than biological or taxonomical preference (like coniferous/deciduous) suggests that being generalists, they are not particular about the nature of their food source. At the same time, the correlation between canopy cover and tree density, despite being statistically significant, is weak at low densities, suggesting that the preference of birds for low density patches might be driven by some factor other than light availability. For instance, Menon and Shahabuddin (2021) found that tree density across various sites in a low-latitude temperate forest, comprising both protected and managed patches, was highly negatively associated with mean tree girth, and observations suggested a negative relationship with canopy height as well. While these could be due to suppressed regeneration and sapling growth, and therefore indirectly due to light availability (shade effects), both variables were positively linked with bird abundance and richness (Menon and Shahabuddin, 2021). Wider and taller trees are expected to host a larger and more diverse group of arthropod prey species as a result of greater diversity of microhabitats. Wolfe et al. (2014) suggested that because forest-dwelling arthropod species are often cryptic (Bond, 2007), it may be inherently difficult and inefficient for an insectivore to directly assess the availability of arthropods, particularly in resource-scarce periods. They suggested this to be the reason why insectivores, unlike frugivores, tracked structural cues. However, our ‘invertebrate-feeding’ species are naturally specialised on cryptic prey species. This could be why they tracked the qualitative features of the food resources (indirectly by the coniferous/deciduous distinction) over structural characteristics. On the other hand, the generalist omnivores would have had difficulty foraging for these cryptic species, and hence attained higher foraging efficiency by tracking structural characteristics, especially with the formation of foraging flocks.

Birds generally show a hump-shaped trend with proportion of deciduous trees, with highest numbers found in mixed forests and lowest in both purely coniferous and purely broadleaved patches (e.g., Alatalo, 1981; Jansson and Andrén, 2003; Hagar, 2007; Ohno and Ishida, 1997; Felton et al., 2010). However, in alignment with Alatalo's findings (1981) that relative use of such deciduous patches were strong in summer but declined in winter, omnivores in our study in the summer-winter transition showed neither a linear nor a polynomial relationship with proportion of deciduous trees, while invertebrate-feeder abundance declined with deciduous proportion. Oaks are usually linked with higher biodiversity (especially *Q. robur*, which is highly insect-rich and has the highest visitation rates by birds), but oak-rich points in our study were poorest for birds in both periods. It is to be noted that a majority of the herbivorous arthropods on oaks belong to Lepidoptera (45% of the total species richness of phytophages on *Q. robur/petraea*; Kennedy and Southwood, 1984), which attain a peak density in spring. The trees see relatively very low lepidopteran attacks after mid-June (Feeny, 1970). When considering all phytophagous insects, there is a second but much smaller peak in abundance in late summer and early autumn of suckers, leaf-miners and gall-formers (Southwood et al., 2004). Thus, the overwhelming preference for oaks by arthropods, and consequently by predatory birds, is likely to be prevalent mainly in spring and summer. Moorman et al. (2007) have shown that lepidopterans are less preferred than coleopterans and hymenopterans. The arthropod communities of oaks and their regulation by predators has rarely been explored in the post-breeding periods of late summer and early autumn. Moreover, the specific invertebrate-feeding species present during this summer-autumn transition period such as treecreepers, robin and goldcrest/firecrest are largely specialists and do not prey on external-feeding herbivores. Treecreepers are bark-feeders, robin is mostly a ground-forager and goldcrest is specialised for spruce (firecrest shows no preference for either conifers or broadleaf trees; Leisler and Thaler, 1982). In fact, even omnivores like the great tit have been shown to preferentially feed on conifers like pines (Eeva et al., 1997).

Surprisingly, the dominant ground layer vegetation had a great effect on bird abundances at the community level. Canopy heterogeneity, understorey density and understorey diversity were not particularly different between the different groups of points, and there were only weak patterns with canopy cover, richness and diversity. However, as observed in a central Swedish forest by Berg (1997), the ground vegetation seems to be related to tree species composition. Specifically, there is a complex interplay of conifer presence and tree density. The ‘Moss’ points (two in total), relative to all other points, had the highest tree densities, very low deciduous proportions (lowest oak numbers) and very high spruce numbers (Figure S7; see also Figure S2 A). Some of the ‘Bare’ points (no dominant ground layer vegetation) also had high tree densities and high numbers of spruce, although others had varied characteristics. Bird abundance in both 'Moss' and 'Bare' points was very low in the beginning and very high in the end. Several points other than 'Moss' and 'Bare', including one ‘Graminoid’ point, had relatively low proportions of deciduous trees but these consisted mostly of pine and not spruce. While the relation of different ground layers with pine was similar to that with spruce, it was not as strong. ‘Rubus’ and ‘Vaccinium’ points generally had mixed canopies with high deciduous proportions and low spruce numbers as well as low tree densities ('Rubus' had the lowest). Such points had intermediate values of both average bird abundance and increase in abundance. Finally, while other points had intermediate oak numbers, 'Graminoid' points had the highest (like the 'grass' vegetation in Berg, 1997) and these were also associated with generally low canopy heterogeneity. These points had the lowest average bird abundance as well as its increase.

>>>>

Heterogeneity has been explored before in habitat selection studies, but the focus has mostly been on the habitat and not the canopy. For example, Freemark and Merriam (1986) observed positive trends of bird richness and abundance with overall habitat heterogeneity (combination of species compositions and forest structure) and forest size, which are expected to correlate. Canopy heterogeneity however, along with canopy cover, affect light availability within the forest, which would result in higher foraging efficiency and thus can be crucial in foraging for invertebrates in autumn when the conspicuous prey are depleted. Heterogeneity could also explain higher omnivore abundance in forest edges. In our study, across multiple points of the same area within the same forest tract, the relation between omnivore abundance and canopy heterogeneity evolved with time, suggesting that these preferences are driven directly or indirectly by the seasonally changing conditions. Canopy cover was not included as a predictor and hence its effect cannot be compared, but heterogeneity might potentially be more important to birds than average cover, as heterogeneous patches provide better light conditions whilst retaining some other desirable properties of the habitat. Patches with low cover, despite having good light conditions, might not have sufficiently good vegetation characteristics. This could be indicating two separable properties of the canopy that often go unnoticed.

Despite the absence of a significant correlation between canopy heterogeneity and tree density, the latter was negatively related to bird abundance (at the community level as well in omnivores). Since a vast majority of bird detections were auditory and since auditory detections are not limited by such temperate forests with poor understorey layers, the possibility of poorer detection rates with higher tree density can be ruled out, though it is unknown whether density also affects overall bird behaviour. In omnivores, the structural preference rather than biological preference (like coniferous/deciduous) suggests that being generalists, they are not particular about the nature of their food source. At the same time, the correlation between canopy cover and tree density, despite being statistically significant, is weak at low densities, suggesting that the preference of birds for low density patches might be driven by some factor other than light availability. For instance, Menon and Shahabuddin (2021) found that tree density across various sites in a low-latitude temperate forest, comprising both protected and managed patches, was highly negatively associated with mean tree girth, and observations suggested a negative relationship with canopy height as well. While these could be due to suppressed regeneration and sapling growth, and therefore indirectly due to light availability (shade effects), both variables were positively linked with bird abundance and richness (Menon and Shahabuddin, 2021). Wider and taller trees are expected to host a larger and more diverse group of arthropod prey species as a result of greater diversity of microhabitats. Wolfe et al. (2014) suggested that because forest-dwelling arthropod species are often cryptic (Bond, 2007), it may be inherently difficult and inefficient for an insectivore to directly assess the availability of arthropods, particularly in resource-scarce periods. They suggested this to be the reason why insectivores, unlike frugivores, tracked structural cues. However, our ‘invertebrate-feeding’ species are naturally specialised on cryptic prey species. This could be why they tracked the qualitative features of the food resources (indirectly by the coniferous/deciduous distinction) over structural characteristics. On the other hand, the generalist omnivores would have had difficulty foraging for these cryptic species, and hence attained higher foraging efficiency by tracking structural characteristics, especially with the formation of foraging flocks.

Birds generally show a hump-shaped trend with proportion of deciduous trees, with highest numbers found in mixed forests and lowest in both purely coniferous and purely broadleaved patches (e.g., Alatalo, 1981; Jansson and Andrén, 2003; Hagar, 2007; Ohno and Ishida, 1997; Felton et al., 2010). However, in alignment with Alatalo's findings (1981) that relative use of such deciduous patches were strong in summer but declined in winter, omnivores in our study in the summer-winter transition showed neither a linear nor a polynomial relationship with proportion of deciduous trees, while invertebrate-feeder abundance declined with deciduous proportion. Oaks are usually linked with higher biodiversity (especially *Q. robur*, which is highly insect-rich and has the highest visitation rates by birds), but oak-rich points in our study were poorest for birds in both periods. It is to be noted that a majority of the herbivorous arthropods on oaks belong to Lepidoptera (45% of the total species richness of phytophages on *Q. robur/petraea*; Kennedy and Southwood, 1984), which attain a peak density in spring. The trees see relatively very low lepidopteran attacks after mid-June (Feeny, 1970). When considering all phytophagous insects, there is a second but much smaller peak in abundance in late summer and early autumn of suckers, leaf-miners and gall-formers (Southwood et al., 2004). Thus, the overwhelming preference for oaks by arthropods, and consequently by predatory birds, is likely to be prevalent mainly in spring and summer. The arthropod communities of oaks and their regulation by predators has rarely been explored in the post-breeding periods of late summer and early autumn. Moorman et al. (2007) have shown that lepidopterans are less preferred than coleopterans and hymenopterans. Moreover, the specific invertebrate-feeding species present during this summer-autumn transition period such as treecreepers, robin and goldcrest/firecrest are largely specialists and do not prey on external-feeding herbivores. Treecreepers are bark-feeders, robin is mostly a ground-forager and goldcrest is specialised for spruce (firecrest shows no preference for either conifers or broadleaf trees; Leisler and Thaler, 1982). In fact, even omnivores like the great tit have been shown to preferentially feed on conifers like pines (Eeva et al., 1997).

Surprisingly, the dominant ground layer vegetation had a great effect on bird abundances at the community level. Canopy heterogeneity, understorey density and understorey diversity were not particularly different between these different groups of points, and there were only weak patterns with canopy cover, richness and diversity. However, as observed in a central Swedish forest by Berg (1997), the ground vegetation seems to be related to tree species composition. Specifically, there is a complex interplay of conifer presence and tree density. The ‘Moss’ points (two in total), relative to all other points, had the highest tree densities, very low deciduous proportions (lowest oak numbers) and very high spruce numbers (Figure S7; see also Figure S2 A). Some of the ‘Bare’ points (no dominant ground layer vegetation) also had high tree densities and high numbers of spruce, although others had varied characteristics. Bird abundance in both 'Moss' and 'Bare' points was very low in the beginning and very high in the end. Several points other than 'Moss' and 'Bare', including one ‘Graminoid’ point, had relatively low proportions of deciduous trees but these consisted mostly of pine and not spruce. While the relation of different ground layers with pine was similar to that with spruce, it was not as strong. ‘Rubus’ and ‘Vaccinium’ points generally had mixed canopies with high deciduous proportions and low spruce numbers as well as low tree densities ('Rubus' had the lowest). Such points had intermediate values of both average bird abundance and increase in abundance. Finally, while other points had intermediate oak numbers, 'Graminoid' points had the highest (like the 'grass' vegetation in Berg, 1997) and these were also associated with generally low canopy heterogeneity. These points had the lowest average bird abundance as well as its increase.
